---
title: "ESM 244 - Lab 1"
author: "Sachi Shiroma"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, echo= TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"), show_col_types = FALSE)
```

# To view the column names in data set type 'names(sf_trees)' in console
# To view information of observations and diff tpyes 'summary(sf_trees)' in console

# Part 1: Wragnling and GGplot review 

**Example** Find counts of observation by 'legal_status' and wrangle a bit
```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% # gets count number by group 'legal_status'
  drop_na(legal_status) %>% # drops any row in the 'legal_status' column
  rename(tree_count = n) %>% # renames column name to 'n'
  relocate(tree_count) %>% # reorders columns and brings the "tree_count' column to # the front (far left)
  slice_max(tree_count, n = 5) %>% # takes top 5 highest number of trees
  arrange(-tree_count) # arranges tree count from highest to lowest 
# if you want to do lowest to highest take out '-' sign
```
Make a graph of the top 5 trees:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + # `fct_reorder(legal_status, tree_count)` reorders the trees on graph from lowest to highest # add `-` sign in front of `tree_count` to do highest to lowest
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal Status', y = 'Tree Count') +
  coord_flip() + # rotate the columns horizontally so they're sideways 
  theme_minimal()
```

**Example 2**: Only going to keep observations where legal status is "permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift- cmd- c to comment/uncomment quickly
```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site', caretaker == 'MTA') # keeps only variables in legal_status column that are `Permitted Site` or `MTA` 
# `|` means and/or `,` means and 
```

**Example 3** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude`, and store as `black_wood_acacia_df`

```{r}
black_wood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% # `str_detect` says in this column look for 'Blackwood Acacia'
  select(legal_status, date, lat = latitude, lon = longitude) # keep only columns of `legal_status`, `date`, `longitude and latitude`
# also change latitude and longitude to lat and lon` 

# Make a little graph of location 
ggplot(data = black_wood_acacia_df, aes(x= lon, y = lat)) +
  geom_point(color = 'darkgreen')
```

**Example 4** Use `tidyr::separate()`
```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into =c('specific', 'spp_common'), sep = ' :: ') # Within Species column, Using ' :: ' separate left side of :: as `specific` column, and right side of :: as `spp column` 
```

**Example 5** Use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_') 
# In a new column, 'id_status', combine 'tree_id' and 'legal_status' with '_COOL_' in between (drops column of tree_id and legal_status)
# You can combine them with just a space in between by`sep = ' '

```

### Part 2: make some maps 

***Step 1***:convert the lat/lon to spatial points, `sf_as_sf()`
```{r}
black_wood_acacia_sf <- black_wood_acacia_df %>% 
  drop_na(lon, lat) %>% 
  st_as_sf(coords = c('lon', 'lat'))

###tell r what the coordinate reference system is 
st_crs(black_wood_acacia_sf) <- 4326

ggplot(data = black_wood_acacia_sf) +
  geom_sf(color = 'darkgreen')+
  theme_minimal()
# You do not need to assign aes because `geom_sf` has assigned aes 
```

***Putting a map underneath the graph above***
Read in the SF shapefile and add to map 

```{r}
sf_map <-read_sf(here('data', 'sf_map', 'tl_2017_06075_roads.shp'))
sf_map_transform <- st_transform(sf_map, 4326)

ggplot(data = sf_map_transform)+
geom_sf()
```

Combine the Maps 
```{r}
ggplot() +
  geom_sf(data = sf_map, #this will be the first or bottom layer of the graph, everything else will be stacked on top 
          size = .1, # Change thickness of lines 
          color = 'darkgrey') +
  geom_sf(data = black_wood_acacia_sf, 
          color = 'red',
          size = 0.5)+
  theme_void()+
  labs(title = 'Blackwood Acacia in SF')
```


### Now an interactive map !

```{r}
tmap_mode('view')

tm_shape(black_wood_acacia_sf)+
  tm_dots()
# tm_shape is equivalent to ggplot() and tm_dots is equivalent to geom_sf
```

